---
title: "Programming Module Workshops"
author: "Kylar Duffy"
format: html
editor: visual
---

```{r}
#| message: false

#| warning: false

# Execute Module 1 R script

source("../code/Module_01.R") # change the filepath/name if required
```

```{r}
#| message: false

#| warning: false

# Execute Module 2 R script

source("../code/Module_02.R") # change the filepath/name if required
```

```{r}
#| message: false

#| warning: false

# Execute Module 3 R script

source("../code/Module_03.R") # change the filepath/name if required#| message: false
```

# Workshop 04. The Developers Toolbox and AI

# Introduce GitHub Copilot and the chattr package

```{r}
# Load tidyverse and palmerpenguins libraries
library(tidyverse)
library(palmerpenguins)
```

```{r}
# Create a ggplot scatter plot using the penguins dataset.
# Plot bill_length_mm on the x-axis and bill_depth_mm on the y-axis.
# Colour points by species and use geom_point with size 3.
# Add a minimal theme and clean labels.
ggplot(data = penguins, aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "Bill Length vs Bill Depth in Penguins",
       x = "Bill Length (mm)",
       y = "Bill Depth (mm)",
       color = "Species")
```

```{r}
# Create a boxplot using the penguins dataset.
# Compare flipper_length_mm across different species faceted by island
# Using a custom colour palette.
boxplot <- ggplot(data = penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot() +
  facet_wrap(~ island) +
  scale_fill_manual(values = c("darkorange", "steelblue", "forestgreen")) +
  theme_minimal() +
  labs(title = "Flipper Length by Species and Island",
       x = "Species",
       y = "Flipper Length (mm)",
       fill = "Species")
boxplot
# Remove the grid lines and title and make axis lines black
boxplot + theme(panel.grid = element_blank(),
                 plot.title = element_blank(),
                 axis.line = element_line(color = "black"))
# Keep everything and add grid lines

```

```{r}
# Create a scatter plot mapping body_mass_g against flipper_length_mm
# Add a linear regression trend line (geom_smooth) for each species
scatter_plot <- ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Body Mass vs Flipper Length in Penguins",
       x = "Flipper Length (mm)",
       y = "Body Mass (g)",
       color = "Species")
scatter_plot
```

# chattr (Controversial Workspace Engine)

```{r}
# Generating an API Key
usethis::edit_r_environ()
```

# Connecting chattr to your model

```{r}
# Configure chattr inside my permanent R profile
usethis::edit_r_profile()
```

```{r}
# Launching and Mapping the User Interface
chattr::chattr_app()
```

```{r}
library(ggplot2)

data(iris)

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()

# Change axis names to Sepal Length and Sepal Width
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  labs(x = "Sepal Length", y = "Sepal Width")
```

# Conditional Logic - filtering your data

# Base R: if and else

```{r}
# Simple example: Checking Sea Surface Temperature (SST)
sst <- 30.2
if (sst > 29.5) {
  print("Warning: Marine heatwave threshold exceeded!")
  } else {
    print("SST remains within baseline parameters.")
  } # Round brackets () for logical condition, curly {} for the action if that condition is true
```

# Tidyverse Alternative: Vectorised Conditionals (dplyr)

```{r}
# The if_else(condition, true_output, false_output) function evaluates an entire column item-by-item
# Load tidyverse (if not already loaded)
library(tidyverse)

# Sample coral data
coral_monitoring <- tibble(
    site = c("Site_A", "Site_B", "Site_C"),
    depth_m = c(12, 35, 8)
  )

# Classify depth using if_else
coral_monitoring <- coral_monitoring %>%
  mutate(zone = if_else(depth_m > 30, "Deep Reef", "Shallow Reef", ))
# mutate () creates a new columns that are functions of existing variables
# %>% used to pass the output of one function or object as the first argument into the next function
```

```{r}
# case_when(): stacking multiple if-else blocks becomes unreadable so case_when() provides a clean, sequential eval layput using formula tildes
coral_monitoring <- coral_monitoring %>% 
  mutate(reef_category = case_when(
    depth_m < 10  ~ "Lagoon / Flats",
    depth_m <= 30 ~ "Crest / Slope",
    depth_m > 30  ~ "Mesophotic / Deep",
    TRUE          ~ "Unclassified" # Catch-all remainder
  ))
```

```{r}
# write an r script that initialises a tibble named marine_stations containing a column for salinity with values 35, 28, 32 and 12, use the pipe operator (%>%) and mutate() combined with case_when() to create a new column named environment_type, classify values below 15 as "Estuarine", values between 15 and 30 as "Brackish", and values above 30 as "Marine", run script and verify the table structure

marine_stations <- marine_stations %>%
  mutate(environment_type = case_when(
    salinity < 15 ~ "Estuarine",
    salinity >= 15 & salinity <= 30 ~ "Brackish",
    salinity > 30 ~ "Marine"
  ))
```

# Automation - Iterating Efficiency

```{r}
# Automation means instructing r to repeat a task multiple timeswithout copy pasting. We will cover the foundation (for-loops) and tidyverse workflow (purrr)

# Base R: for-loops
# a for-loop repeats a code chunk for each element in a designated sequence
# Classic loop anatomy
for (year in 2020:2024) {
  print(paste("Processing climate data for year:", year))
}

# If we want to track real world data, loops can iterate over file directories
# Loop across index sequences
transect_lengths <- c(50, 100, 25, 75)
for (i in seq_along(transect_lengths)) {
  print(paste("Transect number", i, "measures", transect_lengths[i], "metres."))
}
```

# Tidyverse Alternative: Functional Programming (purrr)

```{r}
# For loops require you to manage index variables (i) and explicitly set up empty "containers" to store results, which often leads to verbose code and hidden bugs. The tidyverse handles iteration via the purrr package using map functions. 

# Using a standard for-loop
site_areas <- c(144, 400, 625)
results <- numeric(length(site_areas)) # Must build an empty container first

for(i in seq_along(site_areas)) {
  results[i] <- sqrt(site_areas[i])

# Using the purr equivalent
library(purrr)
# Map directly and define your expected output type explicitly
mapped_results <- map_dbl(site_areas, sqrt)

# Iterate a summary function over split lists of data
iris %>% 
  split(.$Species) %>% 
  map(~summary(.x))
```

# Writing Custom Functions

```{r}
# Function Definition
calculate_coral_mortality <- function(initial_count, surviving_count) {
  
  # Logic safety switch using our conditional tools!
  if (initial_count <= 0) {
    stop("Initial count must be greater than zero.")
  }
  
  mortality_rate <- (initial_count - surviving_count) / initial_count
  return(mortality_rate)
}

# Utilizing your custom function
calculate_coral_mortality(initial_count = 120, surviving_count = 84)

```


